---
title: "Lecture 3: Hypothesis Testing"
author: "Dr Nicolaas Puts and Dr Lucas França"
date: "14 October 2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background:

The dataset consists of a sample of $n = 500$ researchers in the UK whose demographic characteristics and working details are being studied. In this dataset you may find the following variables:

-   age: exact age of the researchers (years)
-   sex: registered sex at birth (1: male, 2: female)
-   edu: highest degree the researchers hold (1: Bachelors, 2: Masters', 3: PhD)
-   work: exact time spent at work on a typical day (hours)
-   meetings: exact time spent on meetings on a typical day (hours)
-   satisfaction: satisfaction with the work life balance (1: not at all satisfied, to 7: absolutely satisfied).

## Installing necessary packages

```{r}
# Just run these commands once!
install.packages('foreign')
install.packages('tidyverse')
```

## Loading packages

```{r}

```

## Import dataset

```{r}
url <- 'https://github.com/fans-stats/intro_to_R_stats/blob/main/datasets/researchers_sample.SAV?raw=true'
```

# Plots

```{r}
```

```{r}
```

```{r}
```

## SPSS Practical Tasks

## Task 1

First, identify the type of each variable in the dataset.

-   age is a 'continuous numerical' variable
-   sex is a 'binary categorical' variable
-   edu is an 'ordinal categorical (3 points)' variable
-   work is a 'continuous numerical' variable
-   meetings is a 'continuous numerical' variable
-   satisfaction is a 'categorical ordinal (7 points)' variable

## Task 2

Use the appropriate descriptive indices to identify potential typos and if so, clean the dataset. Use the space below to keep a record of the typos you found.

```{r}
# Summarise data to find typos
```

```{r}
# Remove typos
```

## Task 3

Fill in the blanks to appropriately describe the data (clean dataset).

```{r}
```

```{r}
# Standard deviation work
```

```{r}
# Standard deviation meetings
```

In the sample, there were 201 (40.2%) women and 299 (59.8%) men. Their age varied between 34.12 and 62.05 years old, with 50 % of the researchers being younger than 55.94 years old. In our sample, the clear majority (75.2 %) of the researchers held a PhD degree. On a typical day, the researchers in our sample reported that they worked between 5.76 and 7.47 hours (mean hours of work = 7.47 , SD = 0.51 ). The time spent in meetings per day is on average 2.90 hours (SD = 0.10 ). With respect to the work-life balance, the median satisfaction in our sample was 5 (min = 1 , max = 7).

## Task 4

In all universities, the academic staff is expected to work on average 7 hours per day. Does our sample support this null hypothesis?

Print the test value for the population mean and the sample mean: $\mu_0$ = 7.00 and $\bar{X}$ = 7.47.\
Compute the standard error se = $\frac{sd}{\sqrt{n}}$ = 0.51/22.36 = 0.0228.

```{r}
```

Compute the 95% CI under the null hypothesis.

```{r}
```

```{r}
```

95% CI: \[6.955 ,7.045\]

Do we reject the null hypothesis?

Our sampled value is 7.5 and it is not included in the 95% CI, therefore we reject the null hypothesis in favour of the alternative hypothesis.

## Task 5

Use the appropriate command to test if the average hours that researchers spend at work is 7 hours. Write the null and the alternative hypotheses and report on the results.

$H_0$: the mean hours of work in our sample equals 7. \
$H_\alpha$: the mean hours of work in our sample is different to 7.

```{r}
```

In our sample, the mean hours of work was 7.47. This was statistically different to 7 hours (95% CI: \[0.425, 0.515\]; t=20.652, df=498, p\<0.001). Therefore, we reject the null hypothesis and infer that, on average, researchers work more than 7 hours per day.

##Task 6

Use the appropriate command to test if the average hours that researchers spend at meetings is 2 hours. Write the null and the alternative hypotheses and report on the results.

$H_0$: the mean hours of meetings in our sample equals 2. $H_\alpha$: the mean hours of meetings in our sample is different to 2.

```{r}
```

In our sample, the average hours of meetings was 2.90. This was significantly different to 2 hours (mean difference=0.901, 95% CI=\[0.892, 0.910\]; t=196.684, df=499, p\<0.001). Therefore, we reject the null hypothesis and infer that, on average, researchers are in meetings for more than 2 hours per day.

## Task 7

Use the appropriate command to test if the proportion of female researchers is equal to that of males. Write the null and the alternative hypotheses and report on the results.

$H_0$: the proportion of females in our sample equals 0.5 (50%). \
$H_\alpha$: the proportion of females in our sample is different to 0.5 (50%).

```{r}
```

In our sample, the proportion of females was 40.2%. This was significantly different to 50% ($\chi^2$ = 19.208, df = 1, $p$ < 0.001). Therefore, we reject the null hypothesis and infer that the proportion of female researchers is smaller than the proportion of male researchers.