---
title: "Lecture 6: Correlation and Regression - The linear Model (Tasks)"
author: "Dr Nicolaas Puts and Dr Lucas França"
date: "11 November 2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background:

The data set consists of a sample of n=75 prisoners whose alcohol use was recorded. The following measures have been recorded:

-	alcohol:  Alcohol Use Disorder Identification Test (AUDIT) score (it takes values from 0 to 18). 
-	ethnic: Prisoner’s ethnicity (1=white, 2=black, 3=other)
-	offtype: Offense type (0=non-violent, 1=violent)
-	age: Age in years 
-	sex: Gender (0=female, 1=male, 2=other)
-	sleep: How good the prisoner sleeps at night (1: good sleep, 2: so and so, 3: bad sleep) 
-	communication: the extent to which the prisoner feels that has good communication with the prison psychologist (1: not at all – to 7: great communication).
-	Memoryloss: Measure of memory loss. Score taking values from 0 to 40.
-	Pure alcohol intake: Grams of pure alcohol intake per week in standard UBE units.  
-	Pure alcohol intake parents: Grams of pure alcohol intake per week in standard UBE units by the subject parents.  


# Task 1: Loading Packages

Load the relevant packages for the following analysis.

```{r}
# You should load packages with the command library(NAMEOFTHEPACKAGE)
library(tidyverse)
library(foreign)
```

# Task 2: Loading data

Find the error in the code chunk below:

```{r}
## Line 53 was missing a comma.

url <- 'https://github.com/fans-stats/intro_to_R_stats/blob/main/datasets/alcohol_intake.sav?raw=true'
dataset <- read.spss(file = url, to.data.frame = TRUE)
dataset <- as_tibble(dataset)
dataset
```

# Task 3: Publication ready figure

Plot the relationship between 'purealcohol' and 'purealcohol_parents' colouring according to 'offtype'. Change the colour palette and the theme.

```{r}
dataset %>% 
  
  ggplot(aes(x = purealcohol, 
             y = purealcohol_parents)) +
    geom_point(aes(colour = offtype), 
               size = 3) + 
    scale_color_manual(values = c('#2166ac', '#b2182b'), 
                       name = 'Off. type') + 
    xlab('Pure alcohol intake in UBE (prisoner)') + 
    ylab('Pure alcohol intake in UBE (parents)') + 
    ggtitle('Relationship between prisoners and their parents\' alcohol intake (in UBE)') + 
    theme_bw()
```


# Task 4: Subsetting data

Restrict data to male offenders only and plot a histogram for 'alcohol'.

```{r}
dataset %>% 
  filter(sex == 'male') %>% 
  
  ggplot(aes(x = alcohol)) + 
    geom_histogram(binwidth = 1) + 
    theme_bw()
```

